---
title: "XML Vignette"
author: "Ryan McGovern and Steve Sortijas"
date: "June 6, 2019"
output: rmarkdown::github_document
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r ImportDeathData, include=FALSE}
#Load all necessary packages
library(XML)
library(httr)
library(tidyverse)

#Create URL character string object
url <- "https://data.cdc.gov/api/views/w9j2-ggv5/rows.xml"

#Use xmlParse function (XML) with GET function (httr) to parse XML data; n.b. not using GET function results in error message indicating content is not XML
doc <- xmlParse(GET(url))

#Create object to identify root node of XML data
root <- xmlRoot(doc)

#Use xmlToDataFrame function (XML) with selected root node to create data frame; n.b. not subsetting the root node results in an error
ds<-xmlToDataFrame(root[[1]])

#All variables are character; convert variables which should be numeric
ds[,4:5]<-sapply(ds[,4:5],as.numeric)

#create the variable for decade
ds <- mutate(ds, decade=as.numeric(substr(year,1,3))*10)
```

# Introduction

## XML Data

In this project we use R to process and conduct exploratory data analysis on a publicly available XML dataset. XML, or Extensible Markup Language, is an open standard markup language that is designed to carry data and to enable compatibility. Unlike HTML, it uses tags which are user-defined and new tags can be created as new data is added without redefining existing tags. XML data is highly portable because it is stored in plain text, using a common standard which is not hardware or software dependent. XML is widely used in the development of document formats and industry data standards. This project uses XML data from Data.gov.

## Processing XML Data in R

In this vignette we will use the "XML" package for parsing and processing XML data. Another package called "xml2" is also available for managing XML and HTML data. Both packages are available in CRAN and xml2 is part of the tidyverse. The XML package is the better documented of the two packages and the list of functions in XML surpasses those than in xml2. For instance the XML package offers functions for transforming XML data to data frames and lists, whereas xml2 does not offer such functionality.

However, XML does have its particularities. Because it is well-documented, troubleshooting is often simpler. For instance, we use the httr package in calling the data because simply using the URL alone for this particular function resulted in an error. Whereas, for the xml2 package, using the URL alone did not result in an error.

Using the XML package we use only 3 functions: xmlParse, xmlRoot, and xmlToDataFrame. As the name would suggest, xmlParse parses XML content and generates an R object representing an XML tree. This particular function uses the internal nodes by default. xmlRoot is then used to identify the root node. By subsetting the root node in an object, we can then apply the xmlToDataSet function to transform the xml dataset to an R data frame.

## Dataset Information

For this vignette we will use a dataset consisting of mortality rates and life expectancy rates at birth by race and sex in the U.S. from 1900 to 2015. This data set is publicly available from the [CDC](https://data.cdc.gov/NCHS/NCHS-Death-rates-and-life-expectancy-at-birth/w9j2-ggv5).  

This dataset consists of 1,044 observations with 5 variables relating to: year, race, sex, mortality, and average life expectancy. The dataset is in "long" form. Year, race, and sex are the groupings which contain values for mortality rates and average life expectancy rates. Year is inclusive of every year from 1900 to 2015. Race is a categorical variable consisting of: All Races, Black, and White. Although it is not explicit in CDC's description of the data, we presume that the All Races category is Black and White combined. Sex is a categorical variable consisting of: Both Sexes, Female, and Male. Mortality rate (mortality), or age-adjusted death rates is a numerical variable representing the number deaths per 100,000. The link above describes in detail how mortality rates are calculated for various historical periods. Life expectancy rate (average_life_expectancy) is a numerical variable denominated in years. Per [CDC](https://www.cdc.gov/nchs/nvss/mortality/hist293.htm.), life expectancy data are available up to 2014 and are not consistently available for the black population before 1968. The average life expectancy variable is the only one which contains NAs.

The first contingency table below displays the number of observations for each numerical variable grouped by the race and sex. This shows that for average life expectancy, each group is missing one observation. The second contingency table confirms that the missing data is for 2015.

```{r contingency tables, echo=FALSE}

#Add 2 variables to check for the presence of NA values for the numerical variables
ds2 <- ds %>% mutate(mortality_yn = ifelse(is.na(mortality),0,1), ale_yn = ifelse(is.na(average_life_expectancy),0,1))

#Create contingency table based counting the number of non-NA values for race and sex groupings
table1 <- ds2 %>% filter(mortality_yn == 1 | ale_yn == 1) %>% group_by(race, sex) %>% summarise(mortN = sum(mortality_yn), aleN = sum(ale_yn))

#Create contingency table showing the year with the NA values
#these should be referencing the ds2 data that contains the variable mortality_yn
table2 <- ds2 %>% filter(ale_yn == 0) %>% group_by(year) %>% summarise(mortN = sum(mortality_yn), aleN = sum(ale_yn))

#Display tables, change column names
knitr::kable(table1, col.names = c("Race", "Sex", "Mortality (n)","Life Expectancy (n)"), caption = "Count of Mortality and Life Expectancy Observations by Group")

knitr::kable(table2, col.names = c("Year","Mortality (n)","Life Expectancy (n)"), caption = "Count of Mortality and Life Expectancy Observations by Year")

```

# Life expectancy

## Average Life Expectency by Gender and Decade 

First we will explore the average life expectancy data. To help simplify our analysis, we will create a new variable called "decade" in order to group the years into their respective decades. We will then use this grouping to compare change in the numerical variables over time. 

As discussed previously, we interpret the categorical value of "All Races" in the "race" variable to mean "black" and "white" combined. So for this first table, we will use only the "All Races" observations to show the average life expectancy rates by decade. This table is grouped by sex to highlight the different life expectancy rates between sexes. The bottom row of the table summarizes the differences in life expectancy rates between males and females during each decade.

Looking at this fluctuation its worth considering external influences. Though outside the scope of this assignment it would be interesting to incorporate what years there was a major conflict occuring with US involvement. Could the spread in the 1970s be becase of the Vietnam war and the following years due to veterans returning with injuries? Did the same things cause the spreads increasing between Males an Females in the 40s and to some extent the 1990s? OR were there other gender specific public events that create this spread? 

```{r tables, echo=FALSE}
#Average life expectancy by decade, filtered by "All races", "Female", and "Male"
avgALEDecade <- ds %>% filter(race=="All Races" & sex %in% c("Female","Male")) %>% group_by(sex,decade) %>% summarise(avgALE = mean(average_life_expectancy, na.rm=TRUE))

#Sex by decade table
avgALESex <- avgALEDecade %>% spread(key = decade, value = avgALE)

#Bind difference between average life expectancies
avgALESexDiff <- bind_rows(avgALESex,avgALESex[avgALESex$sex=="Female",2:13]-avgALESex[avgALESex$sex=="Male",2:13])

#add a value for difference
avgALESexDiff[3,1] <- 'Difference'

knitr::kable(avgALESexDiff, caption = "Average Life Expectancy by Sex by Decade", col.names = c("","1900","1910","1920","1930","1940","1950","1960","1970","1980","1990","2000","2010"))

```

The average life expectancy trends upwards over time for both females and males for every decade. Females have a higher average life expectancy compared with males for every decade. From 1900 to 1970, the gap in life expectancy between the two groups more than doubles, with the peak difference in the 1970s (7.7 years difference). After 1970, The life expectancy gap between the two sexes then trends downward. In the tables below, we will explore the life expectancy data by race.


```{r DecadeDiffRace_function, echo=FALSE}
#this function will create the differance in life expectancy table by a supplied race
DD_Race <- function(char_race) {
  #if we arent given a character stop the function
  if (is.character(char_race) == 0) {stop("Value is not a character")}

#Average life expectancy by decade, filtered by "All races", "Female", and "Male"
avgALEDecade <- ds %>% filter(race==char_race & sex %in% c("Female","Male")) %>% group_by(sex,decade) %>% summarise(avgALE = mean(average_life_expectancy, na.rm=TRUE))

#Sex by decade table
avgALESex <- avgALEDecade %>% spread(key = decade, value = avgALE)

#Bind difference between average life expectancies
avgALESexDiff <- bind_rows(avgALESex,avgALESex[avgALESex$sex=="Female",2:13]-avgALESex[avgALESex$sex=="Male",2:13])

#add a value for difference
avgALESexDiff[3,1] <- 'Difference'

return(avgALESexDiff)
}
```

## Average Life Expectency by Gender and Decade Broken Out by Race

When analyzed by race, the life expectancy trends seen above hold true: the life expectancy gap between females and males grows from 1900 to 1970 and then begins decreasing. What is noticeable in this analysis is the gap in life expectancy between black and white populations, particularly in the early 1900s. The average life expectancy is greater than 15 years for whites (both male and female) compared with blacks. However, over the years life expectancy gap between blacks and whites narrows considerably. In 2010 the life expectancy gap between races is reduced to nearly 4 years for both sexes. (Let's use the last table to summarize the differences by race)

```{r dd race call, echo=FALSE}
#call the function for each race 
Black <- DD_Race('Black')
knitr::kable(Black, caption ='Average Life Expectancy by Sex and Decade: Black')
Black[3,1] = 'Black'

White <-DD_Race('White')
knitr::kable(White, caption ='Average life expectancy by Sex and Decade: White')
White[3,1] = 'White'

all_races <-DD_Race('All Races')
#commented out as this is duplicative
#knitr::kable(all_races, caption ='Average life expectancy by Sex and Decade: All Races')
all_races[3,1] = 'All Races'

#create a table of the differences of gender by rate
gender_race_spread <-rbind(Black[3,], White[3,], all_races[3,])
attributes(gender_race_spread)$names[1] <- 'Race'

#plot this as a barchart?
knitr::kable(gender_race_spread, caption = 'Spread of Life Expectancy by Race')

```

## Visuals

The first plot looks at average life expectancy by sex for each decade of the dataset. This grouped bar graph demonstrates the trend towards higher life expectancy for both sexes as well as the persistent gap between male and female life expectancy, which widens and appears to currently be in a narrowing phase.

```{r AvgLifeExpByDecade Chart, echo=FALSE}
#assuming the "all races" category means black and white combined then  we just use the "all races" data for the 2 graphs below where race is not part of the visualization?

#calculate the average life exp by decade and gender
chart_1_Data <- ds %>% filter(sex %in% 
c("Female", "Male")) %>% group_by(decade, sex) %>% summarise(avgAgeDecade = mean(average_life_expectancy, na.rm = TRUE))

g <- ggplot(chart_1_Data, aes(x = decade, fill = sex, weight=avgAgeDecade))
g + geom_bar(position = "dodge") + xlab("Decade") +ylab("Average Life Expectancy")+scale_fill_discrete(name = "Gender") + ggtitle("Average Life Expectancy by Decade")

```

When stratified by race, the faceted bar graph below helps to see how life expectancy for black men and women have narrowed the a wide gap in life expectancy compared with white males and females since 1900. It also shows how the life expectancy gap between  males and females has, in recent history, been larger for the black population compared with whites.

```{r AvgLifeExpByDecade Chart2, echo=FALSE}
#Question about calculations: does "all races" category mean black and white combined? If so, then should we just use the "all races" data for the 2 graphs below where race is not part of the visualization?

#calculate the average life exp by decade and gender
chart_2_Data <- ds %>% filter(sex %in% 
c("Female", "Male")) %>% group_by(decade, sex, race) %>% summarise(avgAgeDecade = mean(average_life_expectancy, na.rm = TRUE))

g <- ggplot(chart_2_Data, aes(x = decade, fill = sex, weight=avgAgeDecade))
g + geom_bar(position = "dodge") + xlab("Decade") +ylab("Average Life Expectancy")+scale_fill_discrete(name = "Gender") + ggtitle("Average Life Expectancy by Decade and Race") + facet_grid(~race)

```

The boxplot below summarizes the life expectancy data by race and sex for all years of the dataset. In other words, each box represents the distribution of life expectancy for the 115 values for each race-sex grouping. One observation here is that, although as a whole, females have higher life expectancy than males, the median life expectancy for white males is similar to the median life expectancy for black females. Also, the "All Races" life expectancy boxplots largely mirror the boxplots for whites, which potentially shows how heavily weighted the white population factors in the All Races data.

```{r chart3, warning=FALSE, echo=FALSE}
#boxplot for sepal length
chart_3_Data <- ds %>% filter(sex %in% c("Female", "Male"))

g <- ggplot(data = chart_3_Data, aes(x=race, y=average_life_expectancy, fill=sex))
g + geom_boxplot() + scale_x_discrete(name = "Race") + ggtitle("Boxplot of Average Life Expectancy by Race and Gender") + ylab("Average Life Expectancy") + labs(fill = "Gender")
```

The line graph below tracks the difference in average life expectancy between males and females over time stratified by race. For all racial categories, this difference remains relatively small closer to the early part of this dataset. The life expectancy difference then grows considerably until the 1970s with a current downward trend since that point.

```{r chart4, warning = FALSE, echo=FALSE}
#boxplot for sepal length
chart_3_Data <-gender_race_spread[,1:13]
chart_3_Data <- gather(data=chart_3_Data, key='decade', value='average_life_expectancy', '1900':'2010')

g <- ggplot(data = chart_3_Data, aes(x = decade, y = average_life_expectancy))

g + geom_line(aes(group=Race,linetype = Race )) +  ggtitle("Difference in Average Life Expectency of Genders by Race") + ylab("Difference in Average Life Expectancy between Genders") + xlab("Decade")

```

## Mortality rates

Similar to the upward trend in average life expectancy rates, the mortality rates for males and females in all racial categories has considerably decreased over time. As you can see from the faceted scatterplot below, mortality rates for blacks were considerably higher than whites from 1900 to the 1930s. Since the 1940s, although mortality rates for blacks of both sexes remained persistently higher than for whites, the difference is considerably less.   

```{r mortalityGraphs, echo=FALSE}

#Faceted scatterplot chart
ggplot(ds, aes(x=as.numeric(year), y=mortality)) + geom_point(aes(color=race)) + facet_grid(~sex) + labs(x="Decade", y="Mortality Rate (deaths per 100,000)", color="", title="Mortality Rates by Sex and Race, 1900-2015") + scale_x_continuous(breaks = c(1900,1940,1980,2020))

```

